Transitioning Towards Continuous Delivery in the B2B Domain: A Case Study

### 2014-12-07 paper.tex
Added content

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitioning towards Continuous Delivery in the B2B Domain: A Case Study}
%
%TODO: check this
%\titlerunning{Abbreviated title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \inst{1,2}, J{\"u}rgen M{\"u}nch \inst{1}
%Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
}
%TODO: check this
%\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{Department of Computer Science, University of Helsinki, \\
P.O. Box 68, FI-00014 University of Helsinki, Finland
\and
Steeri Oy, Tammasaarenkatu 5, 00180 Helsinki, Finland
\email{olli.rissanen@aalto.fi, juergen.muench@cs.helsinki.fi}
}

@@

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires companies to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction times, ensuring the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article presents a case study that is conducted in a medium-sized software company in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. The company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers also require to perform manual acceptance testing, which slows down production deployments. The benefits found from continuous delivery support the case company in solving many of its business problems. The company can expose the software functionality to the customers from an earlier stage, and better utilize customer feedback.
%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}

To deliver value fast and to cope with the increasingly active business environment, companies have to find solutions that improve efficiency and speed. Agile practices \cite{cockburn2002agile} have increased the ability of software companies to cope with changing customer requirements and changing market needs \cite{dzamashvili2010impact}. To even further increase the efficiency, shortening the feedback cycle enables faster customer feedback.

Continuous delivery is a design practice aiming to shorten the delivery cycles and by automating the deployment process. It is an extension to continuous integration, where the software functionality is deployed frequently to customer environment. In continuous delivery, the software is developed in a way that it is always ready for release. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

This study is an exploratory deductive case study, which explores how continuous delivery can be applied in the case company. While existing studies of applying the development models exist \cite{neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous deliver in the B2B domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous delivery also requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

This study is organized as follows. The first chapter is the introduction at hand. The second chapter summarizes the relevant literature and theories to position the research and to educate the reader on the body of knowledge and where the contributions are intended. The third chapter presents the research design: objectives, context and applied methods. The findings are then presented in the fourth chapter, organized according to the research questions. The fifth chapter interprets the main results. Finally, the sixth chapter summarizes the results of study and answers to the research question, discusses the limitations of the study and introduces further research avenues.

\subsection{Background and Related Work}
Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. Continuous delivery aims to solve the problem of how to deliver an idea to users as quickly as possible.

%release cycles
In an agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attemps to keep the software ready for release at all times during development process \cite{cdbook}. Instead of stopping the development process and creating a build as in an agile process, the software is continuously deployed to customer environment. This doesn't mean that the development cycles in continuous delivery are shorter, but that the development is done in a way that makes the software always ready for release.

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automatic deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. While it necessarily doesn't mean that teams release often, keeping the software in a state where a release can be made instantly is often seen beneficial.

%Barriers in CI->CD transition: stairway
Olsson et al. investigate the transition phase from continuous integration to continuous delivery, and identify barriers that companies need to overcome to achieve the transition \cite{olsson2012climbing}. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

The practice of deploying versions continuously allows for faster customer feedback, ability to learn from customer usage data and to focus on features that produce real value for the customer \cite{olsson2012climbing}. Olsson et al. state that in order to shorten the internal verification loop, different parts in the organization should also be involved, especially the product management as they are the interface to the customer \cite{olsson2012climbing}. They also note that finding a pro-active lead customer is essential to explore and form a new engagement model.

\section{Case study}

\subsection{Objective}

@@

\subsection{Research design}
In this research, the units under the study are two teams and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of these development approaches can be gained. The first product, a marketing automation called Dialog, is used through an extensive user interface. The second product under inspection is CDM, a Master Data Management \cite{loshin2010master} solution running as an integrated background application. The objective of this study can therefore be summarized as follows:

%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines} performed within the CDM and Dialog teams. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data. Data triangulation is implemented by interviewing multiple individuals in different positions of the teams. Methodological triangulation is implemented by collecting documentary data regarding the development process from internal documents, and by including observations made by the author in the analysis.

%Data collection
The interview is a semi-structured interview with a standardized set of open-ended questions, which allows deep exploration of studied objects \cite{runeson2009guidelines}. The interviews are performed once with every interviewee. There are a total of 12 interviewees: 6 in each team. The interviewees in the CDM team consist of 5 software designers and one team leader. In Dialog team, the interviewees consist of 3 software designers, a quality assurance engineer, a manager for commercialization and a team leader. Leading questions are avoided on purpose, and different probing techniques such as "What?"-questions are used. The interviews are performed in the native language of the interviewee if possible, otherwise in English, and are recorded in audio format. The audio files are then transcribed into text.

%Data analysis
In this case study the data analysis is based on template analysis, which is a way of thematically analysing qualitative data \cite{king1998template}. The initial template was first formed by exploring the qualitative data for two themes: development process and deployment of software. Through multiple iterations of the data, multiple subthemes were then added to the two existing themes by further coding the data. Attention was paid to different roles of the interviewees.

@@

%When to use CD: http://www.continuousagile.com/unblock/cd_costs_benefits.html
%You provide an online service: Web, SaaS, PaaS, online gaming, online big data, or mobile apps with Web service back ends. If you provide online services, you need to move to continuous delivery, or your competitors will race ahead of you.
The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the development process of software. Customer aspect consists of the customer interaction process and customer commitment.

@@

%-testing costs could be pushed at least partially to users
The developers perceive automated testing and test environments to be the largest technical task. Building a viable test automation has many challenges with architectural decisions and available workload. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate that a version is valid.

@@

%External components
It is also common for B2B applications to have external components that have to be configured when the software is installed or the APIs to these components changed. The configurations for these external components either have to be manually updated, or automated as well. In the case company, both Dialog and CDM are integrated to multiple external components. Currently upon version change these integrations are manually updated when necessary. In continuous delivery, changes that modify the API and therefore break these integrations might have to be manually updated.

%Different configuration per customer
Both of the case company's products are used in multiple different customer environments. This introduces a problem of managing different configurations per customer environment and software instance. Multiple possible solutions exist for the configurations, however, such as creating individual custom profiles per customer, with different configuration for each profile. This has already been implemented in the case company. The case company still faces a challenge of how to automatically determine which customer profile should be used for each environment upon deployment.

One of the main differences between B2B and B2C domains is the production environment. While companies operating in the B2C domain often own the production environment, in B2B domain it is common to install the software on customer environment. Along with the varying customer specific configurations, there exists varying customer environment specific configurations. The customer owned environment introduces multiple challenges: accessing the environment, transferring the software product to the environment and integrating the software to other components within the environment. These challenges related to transferring the software include firewall configurations, required certifications and user rights. The customers also have to be informed on actions performed in their servers. For certain environments, a VPN connections is required. A VPN connection often has a user limit, which can prevent automatic version deployment.

@@

%UAT
%-UAT is often the norm and is required before anything is moved to production
%-in B2B, continuous deployment straight to production is very risky
The basic deployment pipeline in the case company first includes a deploy to a user acceptance testing server, which is then tested manually by either the team or the customer. Only after the version has been acceptance tested and validated to work properly, can the production version be released. Based on the interview results, continuous deployment to production is seen very risky due to the applications playing a major role in running the customers business. However, continuous deployment to a staging environment is seen as very beneficial. Therefore the deployment procedure has to be designed to allow testing the software in a user acceptance testing environment, and making it possible to manually trigger the deployment to production.

Both of the case company's products are developed with a branching model, where feature branches are first thoroughly developed and then integrated to the master branch. Only the master branch is ever released to the customer. However, with continuous delivery the long-lived feature branches should be changed to short-lived and relatively small feature branches to allow exposing new functionality faster to the customers. The benefit of smaller branches is that while there may still be bugs, they're introduced in a fewer number of commits, and entire large features don't have to be rollbacked. Smaller branches also enable faster feedback. While the small feature branches might be common for companies with a relatively new software products, companies that have been developing products for a long time might be more devoted to the practice of feature branches. However, it is still possible to keep the software releasable with long-lived feature branches by hiding new functionality or by constructing it as a series of releasable small changes \cite{cdbook}.
%If the components are sufficiently small, it is preferable to have a single build for the whole application—giving the fastest feedback of all. \cite{cdbook}

%Pull requests
%Triggering
The software applications in B2B often are large and modular applications, as is the case in the case company. CDM is a software component that is compiled from multiple projects. As the deployment process is currently manually triggered by first releasing a version, a suitable time can be chosen each time. With continuous delivery the deployment process is triggered automatically, and with a modular project each module has to be in the correct state in order to produce a coherent version. The point when a deployment is triggered has to be designed to maintain the integrity of the application.
\begin{aquote}{CDM Developer}
"Pull requests aren't always working as is, due to the nature of CDM. There are dependencies between projects, and occasionally merges are done without the will to deploy. However, if pull requests are to be deployed when merged, a better picture of this process is needed for the developers."
\end{aquote}
In the case company, merging a pull request is not a valid trigger of a deployment, since pull requests are project specific and the software products are composed of multiple projects. Some features require changes to multiple projects, and all of the pull requests have to be merged for the feature to work as intended. However, a good CI tool will handle the dependency management \cite{cdbook}.

%Multiple customer environments
Both of the case company's products are used by multiple customers, each having their own environment. As the deployments are currently done manually, the customers receiving each deployment can be manually chosen.However, with a continuous delivery process whenever a feature or a new release is ready to be delivered, it can either be deployed to a single customer or to every customer. Attention has to be paid when configuring the workflow to allow deploying to an arbitrary customer. If only certain customers accept updates on a continuous basis, it has to be possible to prohibit deploying to every customer at once. This is the case especially in the early stages, where all of the customers projects aren't yet in the state of continuous delivery.

%Versioning
Multiple customer environments affects versioning of the software product. In the case company, each customer has a unique configuration of the product, with possibly different versions of certain components. According to Jan Bosch, in an IES environment only a single version exists: the currently deployed one. Other versions are retired and play no role \cite{bosch2012building}. However, with multiple environments, multiple different versions of the software are necessary at least in the early phase. Eventually the product may be developed in such manner that a single version is sufficient. The build script therefore has to be configured to consider the customer-specific components when compiling the builds for each customer.

@@

Due to increased developer responsibility and varying interval of version updates continuous delivery causes, a team leader express concern that the delivery process complicates tracking when deployments are performed, and when features are finished. This also concerns other parties working in the customer interface, such as sales. While this is a concern for version deployments to the UAT environment, especially if developers perform these deployments at will, can production releases still be negotiated with the customer. Neely and Stolt have also found similar results \cite{neely2013continuous}, which they solved by providing mechanisms by which the parties if need of knowledge regarding release times could track the work in real time.

@@

%Customers have been trained to use a certain version of the product
Some customers of the case company are reluctant towards new releases. One of the reasons for this reluctancy is that new releases occasionally contain new bugs. In the case company, customers using Dialog have been trained to perform certain tasks with a certain user interface. The customer might perform these tasks daily, once every two weeks or even less frequently. If the UI changes often, the customers feel lost and initially take more time to perform the tasks. This causes frustration in the users, and visible changes generally increases the reluctancy customers have towards new versions, unless the changes are significantly improving the user experience.  
\begin{aquote}{Dialog product owner}
"The user interface should be easy to use. Now it's relatively hard to learn. If customers have just learned to perform a task, and we change the UI, the feedback is terrible."
\end{aquote}

%Customer reluctancy towards new versions
While user experience could be improved in the long run, customers might feel frustrated when they have to learn to work with modifications. The interviewees have ideas on how the reluctancy can be reduced. If modifications were made more often, it leads to smaller and less notable modifications which might reduce the customer reluctancy towards new versions. When customers become accustomed to the model of continuous version updates, the reluctant attitude might change. Customers also need to understand the concept of continuous delivery, and that the features are eventually implemented to address customer needs and to improve the user experience in the long term. 
\begin{aquote}{Dialog product owner}
"If something changes, communication with the customer is hard. Even if the customers felt that the version included very good changes, if something visible changes it gets alot of negative feedback".
\end{aquote}

The customers are unaware of the problems related to deployment issues unless the deployment has errors. However, if the version contains many UI changes the customer might feel frustrated by being unable to perform certain actions in the old way: 
\begin{aquote}{Dialog product owner}
"Customers have been doing this for their living - they're getting old, they have their own working histories and they're in the working mode. These persons don't approve alterations in the software, and they're not ready for the constant changes."
\end{aquote}

Listing the changed features in changelog entries is especially important when releases are made more often. While the changes become smaller the faster versions are released, customers become less aware of when the version will be updated and when features have changed. Currently the version deployments are negotiated with the customers, and necessary actions are taken to explain the changed features to the customers in meetings or in e-mails. These actions can include changelogs, but also training sessions or meetings. When the deployments are made more often, discussions regarding version releases may be reduced or even ceased.

%Pilot customer
A way to identify the best practices in continuous delivery is to develop the continuous delivery process with a pilot customer. Pilot customer is a company willing to help the company to quickly learn what works and what needs to be improved. The interviewees expressed a desire to first test the continuous delivery process with a single customer that is willing to receive updates in a continuous manner, since the engagement model inevitably differs from the current model. Therefore the company can train customer interaction with continuous delivery, find the actual reasons for version reluctancy, see how the change in deployment process affects customer interaction and even test if the UAT environment is necessary.

The acceptance testing is performed in varying ways. In Dialog context, some customers require to perform manual acceptance testing in a user acceptance testing environment before the product can be deployed into production. Other customers trust the developers to perform the acceptance testing, and deployment to production can be negotiated without a build necessarily requiring to pass the user acceptance testing phase. In CDM context all of the customers perform manual acceptance testing in the user acceptance testing environment. The technical implementation therefore should make it possible to continuously deploy versions to the user acceptance testing environment, and by the push of a button to the production environment. However, if the versions are deployed to user acceptance testing environment very often, customers might feel encumbered by the amount of required testing. The customers also have to be informed whenever a new version is available to the user acceptance testing environment.

>>>>

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitioning Towards Continuous Delivery in the B2B Domain: A Case Study}
%
%TODO: check this
\titlerunning{Continuous Delivery in B2B Domain}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \inst{1,2}, J{\"u}rgen M{\"u}nch \inst{1}
%Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
}
%TODO: check this
\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{Department of Computer Science, University of Helsinki, \\
P.O. Box 68, FI-00014 University of Helsinki, Finland
\and
Steeri Oy, Tammasaarenkatu 5, 00180 Helsinki, Finland
\email{olli.rissanen@steeri.fi, juergen.muench@cs.helsinki.fi}
}

@@

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires companies to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction and helps to ensure that the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article presents a case study that is conducted in a medium-sized software company operating in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. The company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers require to perform manual acceptance testing, while some are reluctant towards new versions. By utilizing continuous delivery, it is possible for the case company to shorten the feedback cycles, increase the reliability of new versions, and reduce the amount of resources required for deploying and testing new releases.

%s such as long feedback cycles, low reliability in new versions and high amount of resources required for releases.

%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}
To deliver value fast and to cope with the increasingly active business environment, companies have to find solutions that improve efficiency and speed. Agile practices \cite{cockburn2002agile} have increased the ability of software companies to cope with changing customer requirements and changing market needs \cite{dzamashvili2010impact}. To even further increase the efficiency, shortening the feedback cycle enables faster customer feedback. Continuous delivery is a design practice that aims to shorten the delivery cycles by automating the deployment process and frequently deploying the software to customer environment. In continuous delivery, the software is developed in a way that it is always ready for releasing. %While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

This study is an exploratory case study, which explores how continuous delivery can be applied in the case company that operates in the B2B domain. While existing studies of applying the design practice exist \cite{olsson2012climbing, neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous delivery in this domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous delivery also requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

This study is organized as follows. The second chapter summarizes the relevant literature and theories to position the research and to educate the reader on the body of knowledge and where the contributions are intended. The third chapter presents the research design. The findings are then presented in the fourth chapter, organized according to the research questions. The fifth chapter interprets the main results, and discusses the limitations of the study. Finally, the sixth chapter summarizes the results of study and answers to the research question, discusses the limitations of the study and introduces further research avenues.

\section{Background and Related Work}
In an agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attempts to keep the software ready for release at all times during development process \cite{cdbook}. Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. It therefore aims to solve the problem of how to deliver an idea to users as quickly as possible.

%release cycles
 %Instead of stopping the development process and creating a build as in an agile process, the software is continuously deployed to customer environment. This doesn't mean that the development cycles in continuous delivery are shorter, but that the development is done in a way that makes the software always ready for release.

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automatic deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. An essential part of continuous delivery is the deployment pipeline, which is an automated implementation of an application’s build, deploy, test and release process \cite{humble2006deployment}. A deployment pipeline can be loosely defined as a consecutively executed set of validations that a software has to pass before it can be released. Common components of the deployment pipeline are a version control system and an automated test suite. Humble and Farley define the deployment pipeline as a set of stages, which cover the path from the commit stage to a build.

%Barriers in CI->CD transition: stairway
Challenges in adopting continuous delivery have been researched in multiple studies. Olsson et al. investigate the organization evolution path and the transition phase from continuous integration to continuous delivery \cite{olsson2012climbing}.  The authors define continuous delivery as one of the final steps in the organization evolution path. The authors identify barriers that companies need to overcome to achieve the transition. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

One of the largest technical challenges is the test automation required for rapid deployment \cite{cdbook, humble2006deployment}. Neely and Stolt found out that with a continuous flow, Sales and Marketing departments lost the track of when features are released \cite{neely2013continuous}. Implementing the deployment infrastructure also requires knowledge from the development and operations team \cite{humble2006deployment}. Another challenge is to sell the vision and reasoning behind continuous delivery to the executive and management level \cite{neely2013continuous}.

%The practice of deploying versions continuously allows for faster customer feedback, ability to learn from customer usage data and to focus on features that produce real value for the customer \cite{olsson2012climbing}. Olsson et al. state that in order to shorten the internal verification loop, different parts in the organization should also be involved, especially the product management as they are the interface to the customer \cite{olsson2012climbing}. They also note that finding a pro-active lead customer is essential to explore and form a new engagement model.

\section{Case study}
To provide insight into extending the development process towards continuous delivery, the following research questions have been chosen: \newline

@@

\subsubsection{Research design}
In this research, the units under the study are two teams within the case company, and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of the development approach can be gained. The first product, a marketing automation, is used through an extensive user interface. The second product under inspection is a Master Data Management \cite{loshin2010master} solution running as an integrated background application.

%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines}, performed within the two teams under study. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data. Data triangulation is implemented by interviewing multiple individuals in different positions of the teams. Methodological triangulation is implemented by collecting documentary data regarding the development process from internal documents, and by including observations made by the author in the analysis.

%Data collection
The interview is a semi-structured interview with a standardized set of open-ended questions, which allows deep exploration of studied objects \cite{runeson2009guidelines}. The interviews are performed once with every interviewee. There are a total of 12 interviewees: 6 in each team. The interviewees in the first team consist of 5 software designers and one team leader. In the second team, the interviewees consist of 3 software designers, a quality assurance engineer, a manager for commercialization and a team leader. Leading questions are avoided on purpose, and different probing techniques such as "What?"-questions are used. The interviews are performed in the native language of the interviewee if possible, otherwise in English, and are recorded in audio format. The audio files are then transcribed into text.

%Data analysis
The data analysis is based on template analysis, which is a way of thematically analysing qualitative data \cite{king1998template}. The initial template was first formed by exploring the qualitative data for two themes: development process and deployment of software. Through multiple iterations of the data, multiple subthemes were then added to the two existing themes by further coding the data. Attention was paid to different roles of the interviewees.

@@

%When to use CD: http://www.continuousagile.com/unblock/cd_costs_benefits.html
%You provide an online service: Web, SaaS, PaaS, online gaming, online big data, or mobile apps with Web service back ends. If you provide online services, you need to move to continuous delivery, or your competitors will race ahead of you.
This section is structured according to the research questions. The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the development process of software. Customer aspect consists of the customer interaction process and customer commitment.

@@

%-testing costs could be pushed at least partially to users
The developers perceive automated testing and test environments to be the largest technical task. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate that a version is valid.

@@

%External components
It is also common for B2B applications to have external components that have to be configured when the software is installed or the APIs to these components changed. The configurations for these external components either have to be manually updated, or automated as well. One of the main differences between B2B and B2C domains is the production environment. Both of the case company's products are used in multiple different customer environments. This introduces a problem of managing different configurations per customer environment and software instance.
%Different configuration per customer

 %This introduces a problem of managing different configurations per customer environment and software instance. While companies operating in the B2C domain often own the production environment, in B2B domain it is common to install the software on customer environment. The customer owned environment introduces multiple challenges: accessing the environment, transferring the software product to the environment and integrating the software to other components within the environment. The challenges related to transferring the software include firewall configurations, required certifications and user rights. The customers also have to be informed on actions performed in their servers. For certain environments, a VPN connections is required. A VPN connection often has a user limit, which can prevent automatic version deployment.

@@

%UAT
%-UAT is often the norm and is required before anything is moved to production
%-in B2B, continuous deployment straight to production is very risky
The basic deployment pipeline in the case company first includes a deploy to a user acceptance testing server, which is then tested manually by either the team or the customer. Only after the version has been acceptance tested and validated to work properly, can the production version be released. Continuous deployment to production is seen very risky due to the applications playing a major role in running the customers business. %However, continuous deployment to a staging environment is seen as very beneficial. %Therefore the deployment procedure has to be designed to allow testing the software in a user acceptance testing environment, and making it possible to manually trigger the deployment to production.

Both of the case company's products are developed with a branching model, where feature branches are first thoroughly developed and then integrated to the master branch. With continuous delivery the long-lived feature branches should be changed to short-lived and relatively small feature branches to allow exposing new functionality faster to the customers, and receive feedback faster. While the small feature branches might be common for companies with a relatively new software products, companies that have been developing products for a long time might be more devoted to the practice of feature branches. %However, it is still possible to keep the software releasable with long-lived feature branches by hiding new functionality or by constructing it as a series of releasable small changes \cite{cdbook}.
%If the components are sufficiently small, it is preferable to have a single build for the whole application—giving the fastest feedback of all. \cite{cdbook}

%The benefit of smaller branches is that while there may still be bugs, they're introduced in a fewer number of commits, and entire large features don't have to be rollbacked. Smaller branches also enable faster feedback.

%Pull requests
%Triggering
%
The software applications in B2B often are large and modular applications, as is the case in the case company. The point when a deployment is triggered has to be designed to maintain the integrity of the application. As the deployment process is currently manually triggered by first releasing a version, a suitable time can be chosen each time. With continuous delivery the deployment process is triggered automatically, and with a modular project each module has to be in the correct state in order to produce a coherent version.
%\begin{aquote}{Developer}
%"There are dependencies between projects, and occasionally merges are done without the will to deploy. However, if pull requests are to be deployed when merged, a better picture of this process is needed for the developers."
%\end{aquote}
%In the case company, merging a pull request is not a valid trigger of a deployment, since pull requests are project specific and the software products are composed of multiple projects. Some features require changes to multiple projects, and all of the pull requests have to be merged for the feature to work as intended. However, a good CI tool will handle the dependency management \cite{cdbook}.

%Multiple customer environments
Both of the case company's products are used by multiple customers, each having their own environments. As the deployments are currently done manually, the customers receiving each deployment can be manually chosen.However, with a continuous delivery process whenever a feature or a new release is ready to be delivered, it can either be deployed to a single customer or to every customer. %Attention has to be paid when configuring the workflow to allow deploying to an arbitrary customer. If only certain customers accept updates on a continuous basis, it has to be possible to prohibit deploying to every customer at once. %This is the case especially in the early stages, where all of the customers projects aren't yet in the state of continuous delivery.

%Versioning
Multiple customer environments affects versioning of the software product. In the case company, each customer has a unique configuration of the product, with possibly different versions of certain components. According to Jan Bosch, in an Innovation Experiment System environment only a single version exists: the currently deployed one. Other versions are retired and play no role \cite{bosch2012building}. However, with multiple environments, multiple different versions of the software are necessary at least in the early phase. %Eventually the product may be developed in such manner that a single version is sufficient. The build script therefore has to be configured to consider the customer-specific components when compiling the builds for each customer.

@@

Due to increased developer responsibility and varying interval of version updates continuous delivery causes, a team leader expresses concern that the delivery process complicates tracking when deployments are performed, and when features are finished. This also concerns other parties working in the customer interface, such as sales. %While this is a concern for version deployments to the UAT environment, especially if developers perform these deployments at will, can production releases still be negotiated with the customer. Neely and Stolt have also found similar results \cite{neely2013continuous}, which they solved by providing mechanisms by which the parties if need of knowledge regarding release times could track the work in real time.

@@

%Customers have been trained to use a certain version of the product
Some customers of the case company are reluctant towards new releases. One of the reasons for this reluctancy is that new releases occasionally contain new bugs. In the case company, customers have been trained to perform certain tasks with a certain user interface. The customer might perform these tasks daily, once every two weeks or even less frequently. If the UI changes often, the customers feel lost and initially take more time to perform the tasks. This causes frustration in the users, and visible changes generally increases the reluctancy customers have towards new versions, unless the changes are significantly improving the user experience.  
\begin{aquote}{Product owner}
"The user interface should be easy to use. Now it's relatively hard to learn. If customers have just learned to perform a task, and we change the UI, the feedback is terrible."
\end{aquote}

%Customer reluctancy towards new versions
%While user experience could be improved in the long run, customers might feel frustrated when they have to learn to work with modifications. If modifications were made more often, it leads to smaller and less notable modifications which might reduce the customer reluctancy towards new versions. Especially UI changes cause frustration.
%\begin{aquote}{Product owner}
%"If something changes, communication with the customer is hard. Even if the customers felt that the version included very good changes, if something visible changes it gets alot of negative feedback".
%\end{aquote}

%The customers are unaware of the problems related to deployment issues unless the deployment has errors. However, if the version contains many UI changes the customer might feel frustrated by being unable to perform certain actions in the old way: 
%\begin{aquote}{Dialog product owner}
%"Customers have been doing this for their living - they're getting old, they have their own working histories and they're in the working mode. These persons don't approve alterations in the software, and they're not ready for the constant changes."
%\end{aquote}

Listing the changed features in changelog entries is especially important when releases are made more often. While the changes become smaller the faster versions are released, customers become less aware of when the version will be updated and when features have changed. Currently the version deployments are negotiated with the customers, and when the deployments are made more often, discussions regarding version releases may be reduced or even ceased.%Currently the version deployments are negotiated with the customers, and necessary actions are taken to explain the changed features to the customers in meetings or in e-mails. These actions can include changelogs, but also training sessions or meetings.

%Pilot customer
A way to identify the best practices in continuous delivery is to develop the continuous delivery process with a pilot customer. Pilot customer is a company willing to help the company to quickly learn what works and what needs to be improved. The interviewees expressed a desire to first test the continuous delivery process with a single customer that is willing to receive updates in a continuous manner, since the engagement model inevitably differs from the current model.% Therefore the company can train customer interaction with continuous delivery, find the actual reasons for version reluctancy, see how the change in deployment process affects customer interaction and even test if the UAT environment is necessary.

The acceptance testing is performed in varying ways. Some customers require to perform manual acceptance testing before the product can be deployed into production. Other customers trust the developers to perform the acceptance testing. The technical implementation therefore should make it possible to continuously deploy versions to the user acceptance testing environment, and by the push of a button to the production environment. However, if the versions are deployed to user acceptance testing environment very often, customers might feel encumbered by the amount of required testing. The customers also have to be informed whenever a new version is available to the user acceptance testing environment. Customers might be using the software when a new version is deployed, and the deployment process shouldn't interfere with ongoing usage.

%Deployment time and downtime
%A customer might be doing a critical task
 %The production deployment date should be negotiated with the customer. %Therefore the most viable solution is to initially deploy the versions to the UAT environment. To prevent downtime and obstructing the usage of the software, the production deployment date should be negotiated with the customer. Production deployment might cause downtime, and cannot be made during times when critical tasks are performed. The test environments can be continuously updated, and once the customer is satisfied with the software and feels a need for a new version, deployment to production can be triggered.

%In the case company, the production deployment date is often negotiated with and specified by the customer, while the product can be deployed to UAT at will. UAT is also used by the customer to validate the functionality of the software, since the customer has ordered the product to perform certain tasks and fill certain criterias. However, the customers don't validate the versions at the UAT environment unless they are informed that new versions are available, since the UAT is not used for any other purpose than testing.
